---
output: html_document
title: "`r params$html_title`"
author: "Zijian Ni"
date: "`r Sys.Date()`"
params:
    html_title: "foo"
    RData_dir: "foo"
---

```{r, include=F}
####################
# No need to edit this file unless adding new functionalities
####################

load(params$RData_dir) # all the rest are based on analysis_out
```

```{r, include=F}
library(ggplot2)
library(Matrix)
library(cowplot)
library(RColorBrewer)
library(grid)
library(readbitmap)
library(dplyr)
library(SpotClean)
library(S4Vectors)

SEED <- 2021

bw_col <- c("#000000","#e9e9e9")
gw_col <- c("#727272","#e9e9e9")
ts_col <- c("#FF9DC7","#ca8bb1")
bg_col <- c("#D8D8D8","#767171")
names(bw_col) <- c("1","0")
names(gw_col) <- c("1","0")
colScale_bw <- scale_fill_manual(values = bw_col)
colScale_gw <- scale_fill_manual(values = gw_col)

slide_info <- list(slide=analysis_out$bcs)
if(!is.null(analysis_out$images_tibble)){
    slide_info$grob <- analysis_out$images_tibble$grob[[1]]
}
raw_obj <- CreateSlide(analysis_out$raw_matrix,
                       slide_info,
                       gene_cutoff = 0)
```

#### Color by spots labels (tissue or not)

```{r,echo=F, dpi=300, fig.width=18, fig.height=4.5}
plots_labels <- VisualizeLabel(raw_obj, label="tissue", 
                               title=paste0(analysis_out$sample_name,
                                            ": raw image with spots labels"), 
                               legend_title = "Tissue spots?")

plots_labels_bw <- VisualizeLabel(raw_obj, 
                                  label="tissue", 
                                  title=paste0(analysis_out$sample_name,
                                               ": spots labels (b/w)"), 
                                  legend_title = "Tissue spots?") + colScale_bw


if(!is.null(analysis_out$images_tibble)){
    plots_raw <- VisualizeSlide(raw_obj)
}else{
    plots_raw <- plots_labels
}


plot_grid(plotlist = list(plots_raw, plots_labels, plots_labels_bw), ncol = 3)
```

#### Color by UMI counts, non-zero genes, mitochondrial proportions


```{r,echo=F, dpi=300, fig.width=18, fig.height=4.5}
plots_umi_counts <- VisualizeHeatmap(
        raw_obj,value = "sum_umi",viridis = FALSE, 
        legend_range = range(metadata(raw_obj)$slide$sum_umi),
        legend_title = "total UMIs",
        title = paste0(analysis_out$sample_name,": total UMI counts")
    )

plots_nz_gene <- VisualizeHeatmap(
        raw_obj,value = "sum_gene",viridis = FALSE, 
        legend_range = range(metadata(raw_obj)$slide$sum_gene),
        legend_title = "nonzero genes",
        title=paste0(analysis_out$sample_name,": number of non-zero genes")
    )

plots_mt_prop <- VisualizeHeatmap(
        raw_obj,value = "mt_prop",viridis = FALSE, 
        logged = FALSE,
        legend_title ="mitochondrial\nproportion",
        title=paste0(analysis_out$sample_name,": mitochondrial UMI proportions")
    )

plot_grid(plotlist = list(plots_umi_counts, plots_nz_gene, plots_mt_prop), ncol = 3)
```

#### For mixed-species data, color by human counts, mouse counts, human proportion

```{r,echo=F, dpi=300, fig.width=18, fig.height=4.5}
if(!is.null(analysis_out$bcs$sum_human_umi)){
    
    plots_human_counts <- VisualizeHeatmap(
        raw_obj,value = "sum_human_umi",viridis = FALSE, 
        legend_range = range(metadata(raw_obj)$slide$sum_human_umi),
        legend_title = "total human UMIs",
        title = paste0(analysis_out$sample_name,": total human UMI counts")
    )
    
    plots_mouse_counts <- VisualizeHeatmap(
        raw_obj,value = "sum_mouse_umi",viridis = FALSE, 
        legend_range = range(metadata(raw_obj)$slide$sum_mouse_umi),
        legend_title = "total mouse UMIs",
        title = paste0(analysis_out$sample_name,": total mouse UMI counts")
    )
    
    plots_human_prop <- VisualizeHeatmap(
        raw_obj,value = "prop_human",viridis = FALSE, logged = FALSE,
        legend_range = range(metadata(raw_obj)$slide$prop_human),
        legend_title = "human UMI\nproportion",
        title = paste0(analysis_out$sample_name,": human UMI proportions")
    )
    
    plot_grid(plotlist = list(plots_human_counts, plots_mouse_counts, 
                              plots_human_prop), ncol = 3)
    
}
```



#### Color by whether total UMIs are above certain cutoffs

```{r,echo=F, dpi=300, fig.width=11, fig.height=24}
plots_umi_cutoff <- list()
cutoffs <- grep("cutoff",colnames(analysis_out$bcs), value = T)
if(length(cutoffs)>0){
    for(cutoff in cutoffs){
        metadata(raw_obj)$slide$cutoff_temp <- factor(analysis_out$bcs[,cutoff])
        plots_umi_cutoff[[cutoff]] <- 
            VisualizeLabel(raw_obj, 
                           label="cutoff_temp", 
                           title=paste0(analysis_out$sample_name,
                                        ": ",cutoff), 
                           legend_title = paste0("Above ",cutoff,"?"))+
            colScale_bw
    }
    plot_grid(plotlist = plots_umi_cutoff, ncol = 2)
}



```

#### Distribution of sum UMIs in tissue and background

```{r, echo=F, dpi=300, fig.width=6, fig.height=4}
ggplot()+geom_density(data=analysis_out$bcs%>%filter(tissue==1),aes(x=sum_umi), color=ts_col[2], fill=ts_col[1], alpha=0.5, size=1.3)+
    geom_density(data=analysis_out$bcs%>%filter(tissue==0),aes(x=sum_umi), color=bg_col[2],fill=bg_col[1], alpha=0.5, size=1.3)+
  scale_x_continuous(trans="log1p",breaks = c(0,10^(0:6)))+
  labs(title = "total UMI counts distribution", subtitle = "Gray: background. Pink: tissue.")+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```


#### Color by the (unnormalized) top 4 highest expressed gene in tissue spots

```{r,echo=F, dpi=300, fig.width=11, fig.height=8}
plots_top4 <- list()
ts_raw <- analysis_out$raw_matrix[,analysis_out$bcs$tissue==1]
top_gene <- rowMeans(ts_raw) %>% 
    sort(decreasing = T) %>% head(4) %>% names

for(gn in top_gene){
    analysis_out$bcs$top_gene <- analysis_out$raw_matrix[gn,]
    
    plots_top4[[gn]] <- VisualizeHeatmap(
        raw_obj, gn, viridis = FALSE,
        title = paste0(analysis_out$sample_name,
                       ": raw UMI counts of top 4 genes")
    )
}

plot_grid(plotlist = plots_top4, ncol = 2)
```


#### UMAP plots colored by Seurat clusters and spots labels

```{r, echo=F, dpi=300, fig.width=14, fig.height=6}
dp1 <- Seurat::DimPlot(analysis_out$S_obj, reduction = "umap", 
               label = TRUE)+
    ggtitle("Cluster")
dp2 <- Seurat::DimPlot(analysis_out$S_obj, reduction = "umap", 
               label = FALSE, group.by = "bc_label", cols=c("#CCCCCC","#FF7BAC"))+
    ggtitle("Tissue spots?")

plot_grid(plotlist = list(dp1,dp2))
```

#### Visualize Seurat clusters in spatial image

```{r,echo=F, dpi=300}
VisualizeLabel(raw_obj, "clust",
               title = paste0(analysis_out$sample_name,": clusters"),
               subset_barcodes = filter(analysis_out$bcs,tissue==1)$barcode,
               legend_title = "Seurat clusters")

VisualizeLabel(raw_obj, "clust",
               title = paste0(analysis_out$sample_name,": clusters"),
               legend_title = "Seurat clusters")

```

#### Visualize individual clusters

Only show clusters with mixed tissue and background spots (mix score $\in [0.02, 0.98]$)

```{r, echo=F}
cand_clust <- c()
for(cl in unique(analysis_out$bcs$clust)){
    
    # proportion of tissue spots within a cluster
    mix_score <- analysis_out$bcs %>% 
        filter(clust==cl) %>% 
        select(tissue,) %>% 
        unlist %>%
        as.character %>%
        as.integer %>% 
        mean
    #print(mix_score)
    
    if(abs(mix_score-0.5)<=0.48){
        cand_clust <- c(cand_clust, cl)
    }
}

plots_singlecluster <- list()

for(i in seq_along(cand_clust)){
    plots_singlecluster[[i]] <- analysis_out$bcs %>%
        filter(clust==cand_clust[i]) %>%
        ggplot(aes(x=imagecol,y=imagerow, color=tissue, fill=clust)) +
        geom_point(shape = 21, size = 1.75, stroke = 0.5)+
        coord_cartesian(expand=FALSE)+
        scale_color_manual(values = bw_col)+
        xlim(0,max(analysis_out$bcs %>% 
                       select(width)))+
        ylim(max(analysis_out$bcs %>% 
                     select(height)),0)+
        xlab("") +
        ylab("") +
        ggtitle(paste0(analysis_out$sample_name,": cluster ",cand_clust[i]))+
        labs(fill = "Seurat cluster")+
        guides(fill = guide_legend(override.aes = list(size=3)))+
        theme_set(theme_bw(base_size = 10))+
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"),
              axis.text = element_blank(),
              axis.ticks = element_blank())
}

f_height <- 4*ceiling(length(cand_clust)/2)
```

```{r, echo=F, dpi=300, fig.width=11, fig.height=f_height}
if(length(cand_clust)>0){
    plot_grid(plotlist = plots_singlecluster, ncol = 2)
}
```
