---
title: "Simulation for spatially variable (SV) gene identification"
author: "Zijian Ni"
date: "`r Sys.Date()`"
output: html_document
---

### Load packages and datasets

```{r, echo=F, include=F}
library(SpotClean)
library(SummarizedExperiment)
library(ggplot2)
library(Matrix)
library(Seurat)
library(dplyr)
library(viridis)
library(cowplot)
library(reshape2)
library(SoupX)
library(celda)
library(reticulate)

SEED <- 2021
source("~/Google Drive/Hallu/codes/ckgroup/spatial_data/simulation/functions_for_simulation.R")

load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/public_spatial_datasets.RData")
sample_names <- c("LIBD_151507","LIBD_151508","LIBD_151669","LIBD_151670","LIBD_151673","LIBD_151674",
                  "mouse_brain","mouse_kidney","human_breast","human_lymphnode","human_spinalcord")
tif_dir <- ("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/figures/Rplots")

raw_ts_list <- list()
for(dset in names(raw_list)){
    top5000 <- rank(-rowMeans(raw_list[[dset]]), ties.method = "first")<=5000
    gene_tokeep <- rowMeans(raw_list[[dset]])>=0.1 & top5000
    raw_list[[dset]] <- raw_list[[dset]][gene_tokeep,slide_all_list[[dset]]$barcode]
    raw_ts_list[[dset]] <- 
        raw_list[[dset]][,filter(slide_all_list[[dset]], tissue==1)$barcode]
}

gene_list <- lapply(raw_ts_list, rownames)
bc_list <- lapply(raw_ts_list, colnames)
```

### Run SpatialDE on real datasets to identify known DE genes

For the simulation purpose, assume no sequencing depth artifact, so do not regress out total counts. Note: SpatialDE transforms data matrix to Pandas DataFrame, which requires tons of memory. Run one data at a time if no enough memory for storing all 11 datasets.

```{python}
import SpatialDE
import NaiveDE
import numpy as np
import scipy
import pandas as pd
import random


spatialde_out = {}

for dset in r.raw_ts_list.keys():
    print(dset)
    # Load data from R
    dat = r.raw_ts_list[dset].toarray()
    dat = pd.DataFrame(dat)
    dat.index = r.gene_list[dset]
    dat.columns = r.bc_list[dset]
    sample_info = r.slide_all_list[dset]
    sample_info = sample_info[sample_info.tissue=="1"]

    # Transform to normal
    norm_expr = NaiveDE.stabilize(dat)

    # Run SpatialDE test
    random.seed(r.SEED)
    spatialde_out[dset] = SpatialDE.run(sample_info[["imagerow","imagecol"]], norm_expr.T)


```

```{r, echo=F, eval=F}
DE_out <- lapply(py$spatialde_out, py_to_r)
for(i in seq_along(DE_out)){
    rownames(DE_out[[i]]) <- DE_out[[i]]$g
}
save(DE_out, file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/SpatialDE_noregress.RData")

```


### Simulate known DE and EE genes

DE genes: Use top 500 highest expressed significant genes from SpatialDE (qval <= 0.01).

EE genes: sample iid Poisson counts using the average expression of each DE gene

```{r, echo=F, eval=F}
sim_matrix <- list()
set.seed(SEED)
for(i in seq_along(raw_ts_list)){
    
    DE_gene <- DE_out[[i]] %>% filter(qval<=0.01) %>% .$g
    
    DE_mat <- raw_list[[i]][DE_gene,]
    ts_idx <- which(colnames(DE_mat)%in%filter(slide_all_list[[i]],tissue==1)$barcode)
    bg_idx <- which(colnames(DE_mat)%in%filter(slide_all_list[[i]],tissue==0)$barcode)
    top_gene <- rowMeans(DE_mat[,ts_idx]) %>% sort(decreasing = T) %>% head(500) %>% names
    DE_mat <- DE_mat[top_gene,]
    rownames(DE_mat) <- paste0("DE_",rownames(DE_mat))
    
    # Generate EE genes by sampling from the mean
    ave_exp <- rowMeans(DE_mat[,ts_idx])

    EE_ts_mat <- t(vapply(ave_exp,
                          function(x) rpois(length(ts_idx),x),numeric(length(ts_idx))))
    colnames(EE_ts_mat) <- colnames(DE_mat[,ts_idx])
    EE_bg_mat <- DE_mat[,bg_idx]

    EE_mat <- cbind(EE_ts_mat, EE_bg_mat)[,colnames(DE_mat)]

    rownames(EE_mat) <- gsub("DE_","EE_",rownames(EE_mat))
    
    sim_matrix[[i]] <- rbind(DE_mat,EE_mat)
}

names(sim_matrix) <- names(raw_list)

```

### Simulate contaminated data

```{r, echo=F, eval=F}
sim_cont_list <- list()
for(dset in names(sim_matrix)){
    sim_cont_list[[dset]] <- sim_spatial_matrix(
        CreateSlide(sim_matrix[[dset]],slide_all_list[[dset]]),
        seed=SEED
    )
}

save(sim_cont_list, 
     file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_genes.RData")
```


### Perform decontamination using SpotClean, SoupX and DecontX

```{r, echo=F, eval=F}
sim_decont_list <- list(soupx=list(), decontx=list(), SpotClean=list())

for(dset in names(sim_cont_list)){
    
    print(dset)
    
    raw_tmp <- sim_cont_list[[dset]]$obs_mat
    raw_ts_tmp <- raw_tmp[,filter(slide_all_list[[dset]], tissue==1)$barcode]
    
    # SoupX
    sim_decont_list$soupx[[dset]] <- run_soupx(raw_tmp, slide_all_list[[dset]], seed=SEED)

    # DecontX
    set.seed(SEED)
    sim_decont_list$decontx[[dset]] <- decontX(raw_ts_tmp)$decontXcounts
    
    
    sim_obj <- CreateSlide(sim_cont_list[[dset]]$obs_mat,
                           slide_all_list[[dset]],
                           gene_cutoff = 0, verbose = F)
    
    # SpotClean
    set.seed(SEED)
    sim_decont_list$SpotClean[[dset]] <- SpotClean(sim_obj,rownames(sim_obj))

}


save(sim_decont_list, 
     file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_decont.RData")
```

```{r, echo=F}
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_genes.RData")
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_decont.RData")
```

### Check SoupX and DecontX performance in simulated EE genes

```{r, echo=F, dpi=300, fig.width=20, fig.height=4, warning=F}
for(dset in names(sim_cont_list)[6]){
    if(!is.character(sim_decont_list$soupx[[dset]])){
        EE_genes <- grepl("EE_",rownames(sim_cont_list[[dset]]$obs_mat))
        slide_tmp <- slide_all_list[[dset]]
        
        raw_exp <- sim_cont_list[[dset]]$obs_mat[EE_genes,]
        soupx_exp <- sim_decont_list$soupx[[dset]][EE_genes,]
        decontx_exp <- sim_decont_list$decontx[[dset]][EE_genes,]
        spotclean_exp <- assay(sim_decont_list$SpotClean[[dset]])[EE_genes,]
        
        raw_exp <- colSums(raw_exp/apply(raw_exp,1,max))
        soupx_exp <- colSums(soupx_exp/apply(soupx_exp,1,max))
        decontx_exp <- colSums(decontx_exp/apply(decontx_exp,1,max))
        spotclean_exp <- colSums(spotclean_exp/apply(spotclean_exp,1,max))
        
        dp1 <- VisualizeHeatmap(
            slide_tmp,raw_exp,logged = F, viridis = F,
            subset_barcodes = filter(slide_tmp, tissue==1)$barcode,
            title = paste0(dset,"_raw_scale"),legend_title = "sum of\nexpression")#+theme_nothing()
        dp2 <- VisualizeHeatmap(
            filter(slide_tmp,tissue==1),logged = F, viridis = F,soupx_exp,
            title = paste0(dset,"_SoupX_scale"),legend_title = "sum of\nexpression")#+theme_nothing()
        dp3 <- VisualizeHeatmap(
            filter(slide_tmp,tissue==1),logged = F, viridis = F,decontx_exp,
            title = paste0(dset,"_DecontX_scale"),legend_title = "sum of\nexpression")#+theme_nothing()
        dp4 <- VisualizeHeatmap(
            filter(slide_tmp,tissue==1),logged = F, viridis = F,spotclean_exp,
            title = paste0(dset,"_SpotClean_scale"),legend_title = "sum of\nexpression")#+theme_nothing()
        
        # tiff(file.path(tif_dir,paste0(dset,"_simEE_scale_exp_l.tif")),units = "in", res = 300, width = 22, height = 5)
        plot(plot_grid(dp1, dp4, dp2, dp3, ncol=4, align="vh"))
        # dev.off()
    }
}
```

### Run SpatialDE on simulated contaminated data and decontaminated data from SoupX, DecontX and SpotClean

```{r, echo=F, eval=F}


if(is(sim_decont_list$SpotClean[[1]],"SummarizedExperiment")){
    sim_decont_list$SpotClean <- lapply(sim_decont_list$SpotClean, function(x) assay(x))
}

list_toDE <- c(list(raw=lapply(sim_cont_list, function(x) x$obs_mat)),
               sim_decont_list)

for(dset in names(sim_cont_list)){
    if(is.character(list_toDE$soupx[[dset]])){
        list_toDE$soupx[[dset]] <- NULL # soupx failed in some data

    }
    list_toDE$raw[[dset]] <- list_toDE$raw[[dset]][,slide_all_list[[dset]]$tissue==1]
}

gene_list <- lapply(list_toDE$raw, rownames)
bc_list <- lapply(list_toDE$raw, colnames)
```

```{python}
import SpatialDE
import NaiveDE
import numpy as np
import scipy
import pandas as pd
import random

spatialde_out = {}

for decont in r.list_toDE.keys():
    dat_mats = r.list_toDE[decont]
    spatialde_out[decont] = {}
    for dset in dat_mats.keys():

        dat = dat_mats[dset].toarray()
        dat = pd.DataFrame(dat)
        dat.index = r.gene_list[dset]
        dat.columns = r.bc_list[dset]
        sample_info = r.slide_all_list[dset]
        sample_info = sample_info[sample_info.tissue=="1"]

        norm_expr = NaiveDE.stabilize(dat)

        random.seed(r.SEED)
        spatialde_out[decont][dset] = SpatialDE.run(sample_info[["imagerow","imagecol"]], norm_expr.T)

```

```{r, eval=F, echo=F}
py_out <- py$spatialde_out

DE_out <- list()
for(decont in names(py_out)){
    for(dset in names(py_out[[decont]])){
        DE_out[[decont]][[dset]] <- py_to_r(py_out[[decont]][[dset]])
    }
}
save(DE_out, 
     file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_decont_SpatialDE.RData")
```



### Compare tested DE with true DE

```{r, echo=F}
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_DE_EE_decont_SpatialDE.RData")

# q_cutoffs <- c(10^-(10:1),3:20*0.05)
q_cutoffs <- c(0.01, 0.05, 0.1)

ROC_list <- list()

for(decont in names(DE_out)){
    
    ROC_list[[decont]] <- list()
    
    for(dset in names(slide_all_list)){
        if(is.null(DE_out[[decont]][[dset]])){
            ROC_list[[decont]][[dset]] <- NA
        }else{
            ROC_mat <- matrix(0, 4, length(q_cutoffs))
            rownames(ROC_mat) <- c("FDR","FPR","TPR","q_cutoff")
            colnames(ROC_mat) <- paste0("q_",q_cutoffs)
            
            for(i in seq_along(q_cutoffs)){
                
                test_DE <- DE_out[[decont]][[dset]]$qval<=q_cutoffs[i]
                real_DE <- grepl("DE_",DE_out[[decont]][[dset]]$g)
                
                ROC_mat[1,i] <- sum(test_DE & !real_DE)/sum(test_DE) # FDR
                ROC_mat[2,i] <- sum(test_DE & !real_DE)/sum(!real_DE) # FPR
                ROC_mat[3,i] <- sum(test_DE & real_DE)/sum(real_DE) # TPR, power
            }
            ROC_mat[4,] <- q_cutoffs
            
            ROC_list[[decont]][[dset]] <- ROC_mat
        }
    }
}

ROC_list <- purrr::transpose(ROC_list)

ROC_list <- ROC_list[c(6:11,1:5)]
names(ROC_list) <- sample_names

# Leave blank space for Soupx in human_lymphnode
ROC_list$human_lymphnode$soupx <- ROC_list$human_lymphnode$raw
ROC_list$human_lymphnode$soupx[1:3,] <- 0
```


```{r, echo=F, dpi=300, fig.width=8, fig.height=7}
myPalette <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "Spectral")))

FDR_p_list <- bar_p_list <- list()
for(dset in names(ROC_list)){
    dset_df <- bind_rows(lapply(ROC_list[[dset]], function(x) data.frame(t(x))), 
              .id = "decont")
    dset_df$decont <- factor(dset_df$decont, levels = c("raw","SpotClean","soupx","decontx"))
    dset_df <- filter(dset_df,!is.na(FDR))
    
    
    FDR_p_list[[dset]] <- (ggplot(dset_df, aes(y=q_cutoff, x=FDR, color=decont, group=decont))+
                            geom_line(size=1)+labs(title=dset, y="q-value cutoff")+
        theme_set(theme_bw(base_size = 10)) +
        theme(
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.background = element_blank(),
            axis.line = element_line(colour = "black")
        ))
    
    bar_p_list[[dset]] <- ggplot(dset_df, aes(y=FDR,x=factor(q_cutoff),fill=decont))+
        geom_bar(position = "dodge",stat = "identity")+labs(title=dset)+
        ylim(c(0,0.5))+scale_fill_manual(values = rev(viridis(4)))+  # myPalette(4)
        theme(
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.background = element_blank(),
            axis.line = element_line(colour = "black"),
            axis.title = element_blank(),
            legend.position = "none"
        )
    
    
}

# tiff("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/figures/Rplots/sim_SV.tif",
#      res = 300, units = "in", width = 7, height = 6)
plot_grid(plotlist = bar_p_list, ncol = 3, align = "vh")
# dev.off()

```


### Session Information

```{r}
sessionInfo()
```