---
title: "Simulation for MSE evaulation"
author: "Zijian Ni"
date: "`r Sys.Date()`"
output: html_document
---

### Load data and functions

```{r, echo=F}
library(SpotClean)
library(SummarizedExperiment)
library(ggplot2)
library(Matrix)
library(dplyr)
library(viridis)
library(cowplot)
library(reshape2)
library(Seurat)
library(SoupX)
library(celda)

SEED <- 2021

source("~/Google Drive/Hallu/codes/ckgroup/spatial_data/simulation/functions_for_simulation.R")

load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/public_spatial_datasets.RData")
sample_names <- c("LIBD_151507","LIBD_151508","LIBD_151669","LIBD_151670","LIBD_151673","LIBD_151674",
                  "mouse_brain","mouse_kidney","human_breast","human_lymphnode","human_spinalcord")

slide_obj_list <- list()
for(dset in names(raw_list)){
    slide_obj_list[[dset]] <- CreateSlide(raw_list[[dset]], 
                                          list(slide=slide_all_list[[dset]]),
                                          gene_cutoff = 0)
}

names(sample_names) <- names(slide_obj_list)[c(6:11,1:5)]

rm(raw_list)
gc()
```

### Simulation based on mouse_brain, mouse_kidney, breast_cancer, lymph_node, targeted_spinalcord, 6 spatialLIBD datasets

Use top 3000 highly expressed genes in raw data as true expression. 
For illustration purpose, use manually specified parameters for simulation.

Try different kernels: linear, laplace, cauchy

```{r, echo=F}
sim_list <- sim_obj_list <- list()
for(dset in names(slide_obj_list)){
    print(dset)
    sim_list[[dset]] <- sim_spatial_matrix(slide_obj_list[[dset]],
                                           umi_per_spot = 2e4,
                                           kernel = "cauchy",
                                           seed=SEED)

    sim_obj_list[[dset]] <- CreateSlide(sim_list[[dset]]$obs_mat,
                                list(slide=slide_all_list[[dset]],
                                     cont_rate=sim_list[[dset]]$cont_rate),
                                gene_cutoff = 0, verbose = F)
}

```

Compare simulated data with real data. 

```{r, echo=F, dpi=300, fig.width=14, fig.height=6}
for(dset in names(slide_obj_list)){
    
    target <- rowSums(assay(slide_obj_list[[dset]])) %>% 
        sort(decreasing = T) %>%
        head(20) %>% 
        tail(1) %>%
        names

    vs1 <- VisualizeHeatmap(slide_obj_list[[dset]], target,
                           title = paste0(dset,": real_data_exp"))
    vs2 <- VisualizeHeatmap(sim_obj_list[[dset]], target,
                           title = paste0(dset,": sim_data_exp"))
    
    plot(plot_grid(vs1, vs2, ncol = 2, align = "vh"))

}

```

### Perform decontamination using SoupX, DecontX, SpotClean

```{r, echo=F, eval=F}
sim_decont_list <- list(soupx=list(), decontx=list(), SpotClean=list())

for(dset in names(sim_obj_list)){
    
    print(dset)
    
    raw_tmp <- sim_list[[dset]]$obs_mat
    raw_ts_tmp <- raw_tmp[,filter(slide_all_list[[dset]], tissue==1)$barcode]
    
    # SoupX
    sim_decont_list$soupx[[dset]] <- run_soupx(raw_tmp, slide_all_list[[dset]], seed=SEED)

    # DecontX
    set.seed(SEED)
    sim_decont_list$decontx[[dset]] <- decontX(raw_ts_tmp)$decontXcounts
    
    # SpotClean
    set.seed(SEED)
    sim_decont_list$SpotClean[[dset]] <- SpotClean(sim_obj_list[[dset]],rownames(sim_obj_list[[dset]]))

}

save(sim_obj_list,sim_decont_list, 
     file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_decont_forMSE.RData")
```

### Check estimation of contamination rate

Lymph node: probably due to not converging. Increasing max iteration will make it better.

```{r, echo=F, dpi=300, fig.width=7,fig.height=6}
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/sim_decont_forMSE.RData")

gp_list <- list()
for(dset in names(slide_obj_list)[c(6:11,1:5)]){
    print(dset)
    true_rates <- sim_list[[dset]]$cont_rate
    est_rates <- sim_decont_list$SpotClean[[dset]]@metadata$contamination_rate
    df_tmp <- data.frame(true_rates=true_rates,est_rates=est_rates)
    gp <- ggplot(df_tmp, aes(x=true_rates,y=est_rates))+
        geom_point(shape=20)+
        ggtitle(sample_names[dset])+
        geom_abline(slope = 1, color="pink")+lims(x=c(0,1),y=c(0,1))+
        theme_classic()+theme(axis.title = element_blank())
    gp_list <- c(gp_list,list(gp))
    
}

# tiff("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/figures/Rplots/sim_cont_rate.tif",
#      res = 300, units = "in", width = 7, height = 7)
plot_grid(plotlist = gp_list, ncol = 3, align = "vh")
# dev.off()
```

# MSE between decontaminated matrix and true expression matrix

```{r, echo=F}

if(is(sim_decont_list$SpotClean[[1]],"SummarizedExperiment")){
    sim_decont_list$SpotClean <- lapply(sim_decont_list$SpotClean, function(x) assay(x))
}

MSE_mat <- matrix(0, length(sim_decont_list[[1]]),
                  length(sim_decont_list)+1)
rownames(MSE_mat) <- names(sim_decont_list[[1]])
colnames(MSE_mat) <- c("raw", names(sim_decont_list))

for(dset in rownames(MSE_mat)){
    ts_bc <- colnames(sim_decont_list$SpotClean[[dset]])
    obs_ts_mat <- sim_list[[dset]]$obs_mat[,ts_bc]
    
    MSE_mat[dset,"raw"] <- mean((obs_ts_mat-sim_list[[dset]]$true_mat[,ts_bc])^2)
    
    for(method in names(sim_decont_list)){
        if(is.character(sim_decont_list[[method]][[dset]])){
            MSE_mat[dset,method] <- NA
        }else{
            MSE_mat[dset,method] <- mean((sim_decont_list[[method]][[dset]]-
                                              sim_list[[dset]]$true_mat[,ts_bc])^2)
        }
        
    }
}


round(MSE_mat,3)

```




### Session Information

```{r}
sessionInfo()
```