---
title: "Decontamination on SpatialLIBD datasets"
author: "Zijian Ni"
date: "`r Sys.Date()`"
output: html_document
---

### Load packages

```{r, echo=F}
library(SpotClean)
library(ggplot2)
library(Matrix)
library(dplyr)
library(viridis)
library(cowplot)
library(reshape2)
library(SoupX)
library(celda)
library(Seurat)
library(scran)
library(scuttle)
library(spatialLIBD)
SEED <- 2021
source("~/Google Drive/Hallu/codes/ckgroup/spatial_data/simulation/functions_for_simulation.R")

tif_dir <- "~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/figures/Rplots"

myPalette <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "Spectral")))
```

### Load SpatialLIBD data

Raw data came from their Github repo. Download processed data using SpatialLIBD package.

Also load another Visium data V1_Adult_Mouse_Brain for imputation.

```{r, echo=F, eval=F}
# Load another public data slide for imputation
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/V1_Adult_Mouse_Brain/V1_Adult_Mouse_Brain.RData")
slide_known <- analysis_out$bcs


# raw count matrix directory
raw_dir <- "~/Google Drive/Hallu/DATA/10x/Spatial/spatialLIBD"
raw_files <- list.files(raw_dir, "h5", full.names = T)
sample_names <- gsub(".*LIBD/(.*)_raw.*","\\1",raw_files)

# Load processed data
ehub <- ExperimentHub::ExperimentHub()
sce_LIBD <- fetch_data(type = "sce", eh = ehub, 
                  destdir = "~/Google Drive/Hallu/DATA/10x/Spatial/spatialLIBD/")

sce_layer <- fetch_data(type = "sce_layer", eh = ehub)
modeling_results <- fetch_data("modeling_results", eh = ehub)

# Load raw matrix data

raw_matrix <- list()
for(i in seq_along(raw_files)){
    raw_matrix[[sample_names[i]]] <- Read10X_h5(raw_files[i])
}
names(raw_matrix) <- sample_names

# Generate per-sample slides
sce_slide <- sce_LIBD@colData@listData
slide_list <- list()
for(dset in sample_names){
    slide_list[[dset]] <- sce_slide %>% data.frame %>% filter(sample_name==dset) %>% arrange()
}
```

Impute background spots.

```{r, echo=F, eval=F}
slide_list_impute <- list()
for(dset in sample_names){
    slide_impute <- slide_list[[dset]] %>% select(barcode, tissue, row, col, imagerow, imagecol) %>% arrange(row, col)
    bg_slide <- slide_known %>% select(barcode, tissue, row, col, imagerow, imagecol) %>%
        filter(!barcode%in%slide_impute$barcode) %>% arrange(row, col)
    bg_slide$tissue <- 0
    
    # estimate average row distance and column distance
    lm_row <- lm(imagerow~row, data=slide_impute)
    row_dist <- coef(lm_row)[2]
    lm_col <- lm(imagecol~col, data=slide_impute)
    col_dist <- coef(lm_col)[2]
    
    # impute imagerow and imagecol of background spots
    bg_slide$imagerow <- predict(lm_row, data.frame(row=bg_slide$row))
    bg_slide$imagecol <- predict(lm_col, data.frame(col=bg_slide$col))

    slide_list_impute[[dset]] <- rbind(slide_impute, bg_slide)
    
    slide_list_impute[[dset]]$sum_umi <- colSums(raw_matrix[[dset]][,slide_list_impute[[dset]]$barcode])
    slide_list_impute[[dset]]$width <- slide_list[[dset]]$width[1]
    slide_list_impute[[dset]]$height <- slide_list[[dset]]$height[1]
}

# Add layers to imputed slides
for(dset in sample_names){
    slide_list_impute[[dset]] <- merge(
        slide_list_impute[[dset]],
        slide_list[[dset]][,c("barcode","layer_guess")],
        by="barcode", all.x=TRUE)
    
    # Change barcodes to chr
    slide_list[[dset]]$barcode <- as.character(slide_list[[dset]]$barcode)
    slide_list_impute[[dset]]$barcode <- as.character(slide_list_impute[[dset]]$barcode)
}

save(slide_list,slide_list_impute,raw_matrix, file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/spatialLIBD/12brains.RData")
```

### Create slide objects

Analyze 2 replicates in each of the 3 samples.

```{r, echo=F}
# spatialLIBD
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/spatialLIBD/12brains.RData")
sets_tokeep <- c("sp_151507","sp_151508","sp_151669","sp_151670","sp_151673","sp_151674")
names(raw_matrix) <- names(slide_list_impute) <- names(slide_list) <- 
    paste0("sp_",names(raw_matrix))

# build slide object
slide_obj_list <- list()
for(dset in sets_tokeep){
    slide_obj_list[[dset]] <- CreateSlide(raw_matrix[[dset]],slide_list_impute[[dset]])
}

for(dset in sets_tokeep){
    gene_tokeep <- KeepHighGene(raw_matrix[[dset]])
    print(range(rowMeans(raw_matrix[[dset]][gene_tokeep,])))
}
```

### Visualize slides

```{r, echo=F, dpi=300, fig.width=11, fig.height=12}
dp_list <- list()
for(dset in names(slide_obj_list)){
    cols <- myPalette(7)
    brk <- c("Layer1","Layer2","Layer3","Layer4","Layer5","Layer6","WM")
    if(!"Layer1"%in%slide_list[[dset]]$layer_guess){
        cols <- cols[-(1:2)]
        brk <- c("Layer3","Layer4","Layer5","Layer6","WM")
    }
    
    dp_list[[dset]] <- VisualizeLabel(slide_obj_list[[dset]],"layer_guess",title=dset)+
        scale_fill_manual(values = cols,breaks = brk) 
    
    dp <- VisualizeLabel(slide_obj_list[[dset]],"layer_guess")+
        scale_fill_manual(values = cols,breaks = brk)+theme_nothing()

    # tiff(file.path(tif_dir,paste0(dset,"_layers.tif")),
    #      width = 6.4, height = 5.8,units = "in",res = 300)
    # plot(dp)
    # dev.off()
    
}
plot_grid(plotlist = dp_list, ncol = 2, align = "vh")

```

### Check marker gene expression across stripes of slide regions

Markers: SNAP25, GFAP, PCP4

151507, 151508: row>90-col,row<=90-col+10

151669, 151670: col%in%60:69

151673, 151674: row>100-col,row<=100-col+10

```{r, echo=F}
mk_genes <- c("GFAP","MOBP") # "SNAP25","GFAP","PCP4","MOBP"

for(dset in names(slide_obj_list)[1]){
    
    cols <- myPalette(7)
    brk <- c("Layer1","Layer2","Layer3","Layer4","Layer5","Layer6","WM")
    if(!"Layer1"%in%slide_list[[dset]]$layer_guess){
        cols <- cols[-(1:2)]
        brk <- c("Layer3","Layer4","Layer5","Layer6","WM")
    }
    
    slide_stripe <- filter(slide_obj_list[[dset]]@metadata$slide,row>90-col,row<=90-col+10)

    
    for(mk_gene in mk_genes){
        stripe_df <- data.frame(row=sort(unique(slide_stripe$row)),exp=NA,layer=NA)
        layer_borders <- c()
        for(i in seq_len(nrow(stripe_df))){
            
            rw <- stripe_df$row[i]
            
            ly <- table(filter(slide_stripe,row==rw)$layer_guess)
            if(sum(ly)<3){
                stripe_df$layer[i] <- "background"
                bcs <- filter(slide_stripe,row==rw,tissue==0)$barcode
            }else{
                stripe_df$layer[i] <- names(ly)[which.max(ly)]
                bcs <- filter(slide_stripe,row==rw,layer_guess==stripe_df$layer[i])$barcode
            }
            
            stripe_df$exp[i] <- mean(assay(slide_obj_list[[dset]])[mk_gene,bcs])
            
            if(i>1){
                if(stripe_df$layer[i]!=stripe_df$layer[i-1]){
                    layer_borders <- c(layer_borders,i-1.5)
                }
            }
            
        }
        
        fname_prefix <- file.path(tif_dir,"LIBD_marker_dist",paste(dset,mk_gene,sep = "_"))
        
        # tiff(paste0(fname_prefix,"_exp.tif"),
        #      width = 6.4, height = 5.8,units = "in",res = 300)
        # plot(VisualizeHeatmap(slide_obj_list[[dset]],mk_gene))
        # dev.off()
        # 
        # tiff(paste0(fname_prefix,"_exp_stripe.tif"),
        #      width = 6.4, height = 5.8,units = "in",res = 300)
        # plot(VisualizeHeatmap(slide_obj_list[[dset]],mk_gene,
        #                  subset_barcodes = slide_stripe$barcode)+
        #     theme_nothing())
        # dev.off()
        # 
        # tiff(paste0(fname_prefix,"_layer_stripe.tif"),
        #      width = 6.4, height = 5.8,units = "in",res = 300)
        # plot(VisualizeLabel(slide_obj_list[[dset]],"layer_guess",
        #                subset_barcodes = slide_stripe$barcode)+
        #     scale_fill_manual(values = cols,breaks = brk)+
        #     theme_nothing())
        # dev.off()
        
        gp <- ggplot(stripe_df, aes(x=row,y=exp))+geom_point()+
            geom_smooth(span=0.2, se=F,size=1)+xlim(c(0,max(stripe_df$row)))+ylim(c(0,max(stripe_df$exp)))+
            theme_bw()+
            geom_vline(xintercept = layer_borders, color="pink",size=1,linetype="dashed")+
            theme(axis.title = element_blank(),
                  axis.text.x = element_blank(),
                  axis.ticks.x = element_blank(),
                  axis.text.y = element_text(size = 15,angle = 90,hjust = 0.5),
                  panel.grid.minor = element_blank(),
                  panel.grid.major.x = element_blank())

        # tiff(paste0(fname_prefix,"_dist.tif"),
        #      width = 4, height = 2,units = "in",res = 300)
        plot(gp)
        # dev.off()

    } 
}


```


### Perform decontamination using SpotClean, SoupX and DecontX

```{r, echo=F, eval=F}
decont_obj_list <- list()
for(dset in names(slide_obj_list)){
    decont_obj_list[[dset]] <- SpotClean(slide_obj_list[[dset]])
}
save(decont_obj_list, file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/6brains_decont.RData")
```

### Check estimated contamination rates

Heatmap of contamination rates:

```{r, echo=F, dpi=300, fig.width=6, fig.height=4}
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/6brains_decont.RData")

for(dset in names(decont_obj_list)){
    dp <- VisualizeHeatmap(decont_obj_list[[dset]],
                     decont_obj_list[[dset]]@metadata$contamination_rate,
                     logged = FALSE,
                     title=dset,
                     legend_title="contamination rates", 
                     legend_range = c(0,1))
    plot(dp)
}

melt(
    lapply(decont_obj_list, function(x) x@metadata$contamination_rate)
) %>% ggplot(aes(y=value,color=L1)) + geom_boxplot() + labs(color="data")
```

### Visualize decontaminated genes

For better visualization, trim out values less than 1. 
Show first two samples as examples.

GFAP: Layer1+WM
MBP, MOBP, PLP1: WM
SNAP25, LMO4, SYT1!, GRIN1!, NRGN!, SLC17A7!: Layer2-Layer6
PCP4: Layer5
CCK: Layer2, 3, 6
ARPP19, FXYD6: Layer2, 3, 5, 6
SNCG: Layer3, 5

Marker gene list comes from supplementary file of spatialLIBD paper.

```{r, echo=F, dpi=300, fig.width=10, fig.height=5}
layer_marker_genes <- readxl::read_xlsx("~/Google Drive/Hallu/codes/ckgroup/spatial_data/spatialLIBD/spatialLIBD_supplementary.xlsx",sheet = "Table S5")

marker_genes <- c("GFAP","MBP","SNAP25",
                  "PCP4","CCK")

marker_genes <- c("GFAP","MBP","MOBP","PLP1","SNAP25",
                  "LMO4","SYT1","GRIN1","NRGN","SLC17A7",
                  "PCP4","CCK", "ARPP19","FXYD6","SNCG")
for(dset in names(decont_obj_list)[1]){
    dp_list <- list()
    for(gn in marker_genes){
        cat(gn,dset)
        
        slide_tmp <- metadata(decont_obj_list[[dset]])$slide
        
        dp1 <- VisualizeHeatmap(
            slide_obj_list[[dset]],gn,
            subset_barcodes = slide_list[[dset]]$barcode,
            legend_range = c(1,max(assay(decont_obj_list[[dset]])[gn,])))#+theme_nothing()
        dp2 <- VisualizeHeatmap(
            decont_obj_list[[dset]],gn,
            legend_range = c(1,max(assay(decont_obj_list[[dset]])[gn,])))#+theme_nothing()
        
        plot(plot_grid(dp1, dp2, align = "vh", ncol=2))
        dp_list <- c(dp_list,list(dp1, dp2))
    }
    # tiff(file.path(tif_dir,paste0(dset,"_decont_nol.tif")),res = 300, units = "in", width = 13, height = 26)
    # plot(plot_grid(plotlist = dp_list,  ncol = 2, align = "vh"))
    # dev.off()
}
```

Boxplots for marker genes in different layers

```{r, echo=F, dpi=300, fig.width=6, fig.height=3}

marker_genes <- c("GFAP","MOBP","SNAP25","MBP","PLP1",
                  "LMO4","SYT1","GRIN1","NRGN","SLC17A7",
                  "PCP4","CCK", "ARPP19","FXYD6","SNCG")

for(dset in names(decont_obj_list)[1]){
    
    cols <- myPalette(7)
    brk <- c("Layer1","Layer2","Layer3","Layer4","Layer5","Layer6","WM")
    if(!"Layer1"%in%slide_list[[dset]]$layer_guess){
        cols <- cols[-(1:2)]
        brk <- c("Layer3","Layer4","Layer5","Layer6","WM")
    }
    
    ts_slide <- filter(slide_obj_list[[dset]]@metadata$slide, !is.na(layer_guess))
    marker_df <- data.frame(layer=factor(ts_slide$layer_guess, levels = brk))
    
    dp1_list <- dp2_list <- list()
    
    for(gn in marker_genes){
        marker_df$exp_raw <- assay(slide_obj_list[[dset]])[gn,ts_slide$barcode]
        marker_df$exp_decont <- assay(decont_obj_list[[dset]])[gn,ts_slide$barcode]
        
        max_val <- max(marker_df[,-1])
        dp1 <- ggplot(marker_df, aes(y=exp_raw, fill=layer))+
            geom_boxplot()+labs(title=paste0("\n",gn),y=NULL, fill="Layer")+
            lims(y=c(0,max_val))+
            scale_fill_manual(values = cols,breaks = brk)+
            theme(axis.text.x = element_blank(),
                  axis.ticks.x = element_blank(),
                  axis.text.y = element_text(size = 10),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  legend.position = "none")
        dp2 <- ggplot(marker_df, aes(y=exp_decont, fill=layer))+
            geom_boxplot()+labs(title=paste0("\n",gn),y=NULL, fill="Layer")+
            lims(y=c(0,max_val))+
            scale_fill_manual(values = cols,breaks = brk)+
            theme(axis.text.x = element_blank(),
                  axis.ticks.x = element_blank(),
                  axis.text.y = element_text(size = 10),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  legend.position = "none")
        # tiff(file.path(tif_dir,"LIBD_marker_boxplot_label.tif"),res = 300, 
        #      width = 5, height = 5, units = "in")
        plot(plot_grid(dp1, dp2, ncol = 2, align = "vh"))
        # dev.off()
        
        dp1_list[[gn]] <- dp1
        dp2_list[[gn]] <- dp2
        
        
    }
    
    # tiff(file.path(tif_dir,"LIBD_marker_boxplot_raw.tif"),res = 300,
    #      width = 10, height = 7, units = "in")
    plot(plot_grid(plotlist=dp1_list, ncol = 5, align = "vh"))
    # dev.off()

    # tiff(file.path(tif_dir,"LIBD_marker_boxplot_decont.tif"),res = 300,
    #      width = 10, height = 7, units = "in")
    plot(plot_grid(plotlist=dp2_list, ncol = 5, align = "vh"))
    # dev.off()
}

```


### Normalization before DE

```{r, echo=F}
norm_list <- list(raw=list(),decont=list())
for(dset in names(decont_obj_list)){
    print(dset)
    for(meth in c("raw","decont")){
        if(meth=="raw"){
            ts_mat <- assay(slide_obj_list[[dset]])[,slide_list[[dset]]$barcode]
        }else{
            ts_mat <- assay(decont_obj_list[[dset]])
        }
        
        set.seed(SEED)
        sce <- SingleCellExperiment(list(counts=ts_mat))
        clusters <- quickCluster(sce, min.size=200)
        sce <- computeSumFactors(sce, cluster=clusters)
        min_sf <- min(sce$sizeFactor[sce$sizeFactor>0])
        # deal with potential nonpositive size factors
        sce$sizeFactor[sce$sizeFactor<=0] <- min_sf
        sce <- logNormCounts(sce)
    
        norm_list[[meth]][[dset]] <- logcounts(sce)
    }
}

```


### Check DE results of known DE genes

Check WM v.s. Layer6

Known marker genes used: (1) fdr<=1e-5 in SpatialLIBD, (2) decontaminated by SpotClean (recall that SpotClean filter out low-expressed and low-variable genes), (3) within top 100 highest expressed genes in raw data. 

Then show top 10 genes sorted by fold change.

```{r, echo=F, dpi=300, fig.width=6, fig.height=4}
pairwise_marker_genes <- readxl::read_xlsx("~/Google Drive/Hallu/codes/ckgroup/spatial_data/spatialLIBD/spatialLIBD_supplementary.xlsx",sheet = "Table S4C")

layer_markers <- pairwise_marker_genes %>% 
    filter(`fdr_WM-Layer6`<=1e-5) %>% .$gene

DE_all <- list()
for(dset in names(slide_obj_list)){ # no layer1 in the middle two samples
    cat("\n")
    cat(dset)
    
    # Choose marker genes: small fdr, decontaminated, top expressed, top FC
    
    markers <- layer_markers[
        layer_markers%in%rownames(norm_list$raw[[dset]])
    ]
    markers <- markers[markers%in%metadata(decont_obj_list[[dset]])$decontaminated_genes]
    
    ave_exp <- rowMeans(raw_matrix[[dset]])
    top100 <- ave_exp %>% sort(decreasing = T) %>% head(100) %>% names
    markers <- markers[markers%in%top100]
    
    # spots in different layers
    spot1 <- slide_list[[dset]] %>% filter(layer_guess=="WM") %>% .$barcode
    spot2 <- slide_list[[dset]] %>% filter(layer_guess=="Layer6") %>% .$barcode
    
    # Report top10 DE genes sorted by |logFC|
    logFC <- log(rowMeans(norm_list$raw[[dset]][markers,spot1])/
        rowMeans(norm_list$raw[[dset]][markers,spot2]))
    
    markers <- abs(logFC) %>% sort(decreasing = T) %>% head(10) %>% names
    
    # initialization
    DE_all[[dset]] <- data.frame(gene=markers)
    for(meth in names(norm_list)){
        DE_all[[dset]][[paste0(meth,"_-log10_pval")]] <- NA
        DE_all[[dset]][[paste0(meth,"_logfc")]] <- NA
    }
    # loop for marker genes
    for(i in seq_along(markers)){
        
        gene <- markers[i]
        
        for(meth in names(norm_list)){
            # expressions in two layers
            exp1 <- norm_list[[meth]][[dset]][gene,spot1]
            exp2 <- norm_list[[meth]][[dset]][gene,spot2]
            logfc <- log(mean(exp1)/mean(exp2))
            
            # t-test
            tt <- t.test(exp1, exp2)
            
            DE_all[[dset]][i,paste0(meth,"_-log10_pval")] <- 
                -log10(tt$p.value)
            DE_all[[dset]][i,paste0(meth,"_logfc")] <- logfc
        }
    }
    
    # remove genes not present
    DE_all[[dset]] <- DE_all[[dset]][!apply(DE_all[[dset]],1, function(x) any(is.na(x))),]
    
}

# Visualize fold change and p-value

for(dset in names(DE_all)){
    pl <- plot_grid(
        ggplot(reshape2::melt(DE_all[[dset]],id.vars="gene",
                              measure.vars=paste0(names(norm_list),"_logfc")), 
               aes(y=value, x=gene, fill=variable))+
            geom_bar(position="dodge", stat="identity")+
            theme(axis.text.x = element_text(angle = 45, hjust = 1),
                  axis.title.x = element_blank(),
                  legend.position = "none",
                  panel.grid.minor = element_blank(),
                  axis.title.y = element_text(size = 15),
                  axis.text = element_text(size=10))+
            labs(y="log FC"),
        
        ggplot(reshape2::melt(DE_all[[dset]],id.vars="gene",
                              measure.vars=paste0(names(norm_list),"_-log10_pval")), 
               aes(y=value, x=gene, fill=variable))+
            geom_bar(position="dodge", stat="identity")+
            theme(axis.text.x = element_text(angle = 45, hjust = 1),
                  axis.title.x = element_blank(),
                  legend.position = "none",
                  panel.grid.minor = element_blank(),
                  axis.title.y = element_text(size = 15),
                  axis.text = element_text(size=10))+
            labs(y="-log10 p-value"),
        ncol = 1,align = "vh")
    
    # tiff(file.path(tif_dir,paste0(dset,"_DE.tif")),res = 300, units = "in", width = 5,height = 4)
    plot(pl)
    # dev.off()
}
```

### Review: check new DE and new EE genes after decontamination

```{r, echo=F}
# make use of Seurat FindMarkers() for a more efficient computation

DE_list <- list(raw=list(), decont=list())
for(meth in names(norm_list)){
    for(dset in names(norm_list[[meth]])){
        set.seed(SEED)
        S_tmp <- CreateSeuratObject(norm_list[[meth]][[dset]])
        S_tmp$layer_guess <- as.character(slide_list[[dset]]$layer_guess)
        Idents(S_tmp) <- "layer_guess"
        DE_list[[meth]][[dset]] <- 
            FindMarkers(S_tmp, ident.1 = "WM", ident.2 = "Layer6", test.use = "t", logfc.threshold = 0) %>% 
            mutate(gene=rownames(.))
    }
}

# make sure the number of genes is the same between raw and decont before adjusting p-value
lapply(DE_list$raw, nrow)
lapply(DE_list$decont, nrow)

padj_cutoff <- 0.01
new_DE_list <- new_EE_list <- list()
for(dset in names(norm_list$raw)){

    new_EE <- intersect(filter(DE_list$raw[[dset]],p_val_adj<=padj_cutoff)$gene,
                        filter(DE_list$decont[[dset]],p_val_adj>padj_cutoff)$gene)
    new_DE <- intersect(filter(DE_list$raw[[dset]],p_val_adj>padj_cutoff)$gene,
                        filter(DE_list$decont[[dset]],p_val_adj<=padj_cutoff)$gene)
    
    new_DE_list[[dset]] <- data.frame(gene=new_DE, 
                                      raw_padj=DE_list$raw[[dset]][new_DE,]$padj, 
                                      decont_padj=DE_list$decont[[dset]][new_DE,]$padj)
    new_EE_list[[dset]] <- data.frame(gene=new_EE, 
                                      raw_padj=DE_list$raw[[dset]][new_EE,]$padj, 
                                      decont_padj=DE_list$decont[[dset]][new_EE,]$padj)
}
```


### Session information

```{r, echo=F}
sessionInfo()
```