---
title: "Analyzing the mixed-species data"
author: "Zijian Ni"
date: "`r Sys.Date()`"
output: html_document
---

- (HM-1) 2_S2: Mouse duodenum as sausage casing to human punch

- (HM-2) 1_S1: Mouse colon as sausage casing to human punch

- (HM-3) 3_S3: Mouse heart enwrapped in human skin

```{r, echo=F}
library(SpotClean)
library(ggplot2)
library(Matrix)
library(scuttle)
library(Seurat)
library(dplyr)
library(reshape2)
library(cowplot)

SEED <- 2021

rdata_dir <- c("~/Google Drive/Hallu/codes/ckgroup/spatial_data/1_S1_manual/1_S1_manual.RData",
               "~/Google Drive/Hallu/codes/ckgroup/spatial_data/2_S2_manual/2_S2_manual.RData",
               "~/Google Drive/Hallu/codes/ckgroup/spatial_data/3_S3_manual/3_S3_manual.RData")
species_refined_dir <- c("~/Google Drive/Hallu/DATA/10x/Spatial/mix_species/1_S1_manual/species_refined2.csv",
                 "~/Google Drive/Hallu/DATA/10x/Spatial/mix_species/2_S2_manual/species_refined2.csv",
                 "~/Google Drive/Hallu/DATA/10x/Spatial/mix_species/3_S3_manual/species_refined2.csv")
tif_dir <- "~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/figures/Rplots"

myPalette <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "Spectral")))

.theme_void <- function (base_size = 11, base_family = "", base_line_size = base_size/22, 
    base_rect_size = base_size/22) 
{
    half_line <- base_size/2
    t <- theme(line = element_blank(), rect = element_blank(), 
        text = element_text(family = base_family, face = "plain", 
            colour = "black", size = base_size, lineheight = 0.9, 
            hjust = 0.5, vjust = 0.5, angle = 0, margin = margin(), 
            debug = FALSE), axis.text = element_blank(), axis.title = element_blank(), 
        axis.ticks.length = unit(0, "pt"), axis.ticks.length.x = NULL, 
        axis.ticks.length.x.top = NULL, axis.ticks.length.x.bottom = NULL, 
        axis.ticks.length.y = NULL, axis.ticks.length.y.left = NULL, 
        axis.ticks.length.y.right = NULL, legend.box = NULL, 
        legend.position = "none", 
        strip.text = element_text(size = rel(0.8)), strip.switch.pad.grid = unit(half_line/2, 
            "pt"), strip.switch.pad.wrap = unit(half_line/2, 
            "pt"), panel.ontop = FALSE, panel.spacing = unit(half_line, 
            "pt"), plot.margin = unit(c(0, 0, 0, 0), "lines"), 
        plot.title = element_text(size = rel(1.2), hjust = 0, 
            vjust = 1, margin = margin(t = half_line)), plot.title.position = "panel", 
        plot.subtitle = element_text(hjust = 0, vjust = 1, margin = margin(t = half_line)), 
        plot.caption = element_text(size = rel(0.8), hjust = 1, 
            vjust = 1, margin = margin(t = half_line)), plot.caption.position = "panel", 
        plot.tag = element_text(size = rel(1.2), hjust = 0.5, 
            vjust = 0.5), plot.tag.position = "topleft", 
        complete = TRUE)
    ggplot2:::ggplot_global$theme_all_null %+replace% t 
}

find_border <- function(slide, is_tissue=TRUE){
    # Find border spots. i.e. tissue spots adjacent to background spots, or
    # background spots adjacent to tissue spots.
    # Args:
    #   slide (dataframe): the slide data frame
    #   is_tissue (logical): whether return tissue border or background border
    # Returns:
    #   (dataframe) slide dataframe with additional column indicating
    #       border (1) or not (0)
    is_tissue <- as.integer(is_tissue)
    target_positions <- slide %>% filter(tissue==is_tissue) %>% select(row, col)
    remaining_positions_1d <- slide %>% filter(tissue==(1-is_tissue)) %>% select(row, col) %>%
        apply(1, paste, collapse="_")

    is_border <- numeric(nrow(target_positions))
    for(ts_idx in seq_len(nrow(target_positions))){
        sur_spots_1d <- .surrounding_spots(target_positions[ts_idx,]) %>%
            apply(1, paste, collapse="_")

        if(any(sur_spots_1d%in%remaining_positions_1d)){
            is_border[ts_idx] <- 1
        }
    }

    slide$border <- 0
    slide$border[slide$tissue==is_tissue] <- is_border
    slide$border <- as.factor(slide$border)
    return(slide)
}

.surrounding_spots <- function(spot_pos){
    # Return the row/col positions of the 8 surrounding spots of a given spot position in 10x Visium
    # Note: It might give invalid spots if input spot is on the edge of the slide
    # Args:
    #   pos (vector of int): length two vector telling the row and column of input barcode
    # Returns:
    #   (dataframe of int): a 8 by 2 dataframe. Each row is the position of one surrounding spot

    pos_shifts <- rbind(c(-1,-1,0,0,1,1),
                        c(-1,1,-2,2,-1,1))

    sur_pos <- t(as.numeric(spot_pos)+pos_shifts)
    colnames(sur_pos) <- c("row", "col")

    return(sur_pos)
}
```

```{r, echo=F}
sample_names <- gsub(".*manual\\/(.*)_manual.*","\\1",rdata_dir)

slide_list <- slide_obj_list <- down_matrix_list <- top1000_list <- list()

for(i in seq_along(rdata_dir)){
    load(rdata_dir[i])
    species_refined_tmp <- read.csv(species_refined_dir[i])
    species_df <- data.frame(barcode=species_refined_tmp$Barcode,
                             species=species_refined_tmp$species_refined2)

    slide_tmp <- merge(analysis_out$bcs,species_df, by="barcode",all.x=TRUE)
    slide_tmp$species[is.na(slide_tmp$species)] <- "background"
    slide_tmp <- find_border(slide_tmp)
    slide_list[[sample_names[i]]] <- slide_tmp
    
    
    # Remove mixture tissue spots
    is_mixture <- grepl("mixture_",slide_tmp$species)
    slide_tmp <- slide_tmp[!is_mixture,]
    analysis_out$raw_matrix <- analysis_out$raw_matrix[,slide_tmp$barcode]
    
    # Remove genes whose average expression in the correct species are lower 
    # than that in the wrong species

    human_gene <- grepl("GRCh38_",rownames(analysis_out$raw_matrix))
    human_mat <- analysis_out$raw_matrix[human_gene,]
    mouse_mat <- analysis_out$raw_matrix[!human_gene,]
    
    human_spots <- filter(slide_tmp,species=="human")$barcode
    mouse_spots <- filter(slide_tmp,species=="mouse")$barcode
    
    bad_human_gene <- which(
        rowMeans(human_mat[,human_spots])<rowMeans(human_mat[,mouse_spots])
    )
    bad_mouse_gene <- which(
        rowMeans(mouse_mat[,mouse_spots])<rowMeans(mouse_mat[,human_spots])
    )
    
    human_mat <- human_mat[-bad_human_gene,]
    mouse_mat <- mouse_mat[-bad_mouse_gene,]
    
    # Downscale mouse genes
    down_ratio <- sum(human_mat)/sum(mouse_mat)
    mouse_mat <- mouse_mat*down_ratio
    
    # Filter out genes with average expression <0.01
    mouse_mat <- mouse_mat[rowMeans(mouse_mat)>=0.01,]
    human_mat <- human_mat[rowMeans(human_mat)>=0.01,]
    
    down_mat <- rbind(mouse_mat,human_mat)
    
    slide_tmp$sum_umi <- colSums(down_mat)
    slide_tmp$sum_mouse_umi <- colSums(mouse_mat)
    slide_tmp$sum_human_umi <- colSums(human_mat)
    
    slide_obj_list[[sample_names[i]]] <- 
        CreateSlide(down_mat, 
                    list(slide=slide_tmp,
                         grob=analysis_out$images_tibble$grob[[1]]),gene_cutoff = 0)
    
}

# sort human and mouse genes separately
top1000_list <- list()
for(dset in names(slide_obj_list)){
    top1000_list[[dset]] <- list()
    ave_exp <- rowMeans(assay(slide_obj_list[[dset]]))
    human_gene <- grepl("GRCh38",rownames(slide_obj_list[[dset]]))
    top1000_list[[dset]]$human <- ave_exp[human_gene] %>% 
        sort(decreasing=T) %>% head(1000) %>% names
    top1000_list[[dset]]$mouse <- ave_exp[!human_gene] %>% 
        sort(decreasing=T) %>% head(1000) %>% names
    
}

```


### Visualize species labels

```{r, echo=F, dpi=300, fig.width=14, fig.height=6}
for(dset in names(slide_obj_list)){
    cols <- myPalette(7)[c(2,5,7)]
    names(cols) <- c("human","mixture","mouse")
    
    slide_tmp <- slide_list[[dset]]
    slide_tmp$species[grepl("mixture_",slide_tmp$species)] <- "mixture"
    
    # tiff(file.path(tif_dir,paste0(dset,"_species.tif")),
    #      width = 5.3, height = 4.8,units = "in",res = 300)
    dp <- VisualizeLabel(slide_tmp,"species",title=NULL,
                         subset_barcodes = filter(slide_list[[dset]],tissue==1)$barcode)+
        scale_fill_manual(values=cols)+.theme_void()
    plot(dp)
    # dev.off()
}
```

### Visualize human and mouse counts across slide

```{r, echo=F, dpi=300, fig.width=5, fig.height=4}
for(dset in names(slide_obj_list)){
    
    inner_bcs <- filter(slide_obj_list[[dset]]@metadata$slide,
                      border==0)$barcode
    
    # tiff(file.path(tif_dir,paste0(dset,"_human_counts_l.tif")),
    #      width = 5, height = 4.8,units = "in",res = 300)
    dp <- VisualizeHeatmap(slide_obj_list[[dset]],"sum_human_umi",title=NULL, 
                           legend_title = "total human UMIs",
                           subset_barcodes = inner_bcs,
                           viridis = F)#+.theme_void()

    plot(dp)
    # dev.off()
    
    # tiff(file.path(tif_dir,paste0(dset,"_mouse_counts_l.tif")),
    #      width = 5, height = 4.8,units = "in",res = 300)
    dp <- VisualizeHeatmap(slide_obj_list[[dset]],"sum_mouse_umi",title=NULL,
                           legend_title = "total mouse UMIs",
                           subset_barcodes = inner_bcs,
                           viridis = F)#+.theme_void()

    plot(dp)
    # dev.off()
}
```


### Check proportion of human and mouse UMI counts in each spot

```{r, echo=F}
spot_prop_list <- gene_prop_list <- list()

wrong_prop <- matrix(0, 2, length(slide_obj_list))
rownames(wrong_prop) <- c("human","mouse")
colnames(wrong_prop) <- names(slide_obj_list)

LPSS <- BG <- numeric(length(slide_obj_list))
names(LPSS) <- names(BG) <- names(slide_obj_list)

for(dset in names(slide_obj_list)){
    
    #for raw data
    human_gene <- grepl("GRCh38",rownames(slide_obj_list[[dset]]))
    
    raw_mat <- assay(slide_obj_list[[dset]])
    slide_tmp <- metadata(slide_obj_list[[dset]])$slide
    
    human_spots <- filter(slide_tmp,species=="human")$barcode
    mouse_spots <- filter(slide_tmp,species=="mouse")$barcode
    
    # proportion of human and mouse UMIs in each spot
    spot_human_prop <- slide_tmp$sum_human_umi/slide_tmp$sum_umi
    spot_human_prop2 <- slide_tmp$sum_human_umi/sum(raw_mat)
    spot_mouse_prop2 <- slide_tmp$sum_mouse_umi/sum(raw_mat)
    
    spot_wrong_prop <- numeric(nrow(slide_tmp))
    spot_wrong_prop[slide_tmp$species=="background"] <- 1
    spot_wrong_prop[slide_tmp$species=="human"] <- 1-spot_human_prop[slide_tmp$species=="human"]
    spot_wrong_prop[slide_tmp$species=="mouse"] <- spot_human_prop[slide_tmp$species=="mouse"]
    
    # LPSS per-spot
    # Human spot: mouse UMIs / total UMIs
    # Mouse spot: human UMIs / total UMIs
    # Background spot: all UMIs / total UMIs
    lpss <- numeric(nrow(slide_tmp))
    lpss[slide_tmp$species=="background"] <- slide_tmp$sum_umi[slide_tmp$species=="background"]/sum(raw_mat)
    lpss[slide_tmp$species=="human"] <- slide_tmp$sum_mouse_umi[slide_tmp$species=="human"]/sum(raw_mat)
    lpss[slide_tmp$species=="mouse"] <- slide_tmp$sum_human_umi[slide_tmp$species=="mouse"]/sum(raw_mat)
    
    
    spot_prop_list[[dset]] <- 
        data.frame(barcode=slide_tmp$barcode,
                   species=slide_tmp$species,
                   total_counts=slide_tmp$sum_umi,
                   human_counts=slide_tmp$sum_human_umi,
                   mouse_counts=slide_tmp$sum_mouse_umi,
                   human_prop=spot_human_prop,
                   mouse_prop=1-spot_human_prop,
                   human_prop2=spot_human_prop2,
                   mouse_prop2=spot_mouse_prop2,
                   lpss=lpss,
                   wrong_prop=spot_wrong_prop)
    
    
    # proportion of UMIs in human spots for each gene
    gene_human_prop <- rowSums(raw_mat[,human_spots])/rowSums(raw_mat)
    
    # proportion of UMIs in mouse spots for each gene
    gene_mouse_prop <- rowSums(raw_mat[,mouse_spots])/rowSums(raw_mat)
    
    gene_prop_list[[dset]] <- data.frame(gene=rownames(raw_mat),
                                         ave_exp=rowMeans(raw_mat),
                                         species=ifelse(human_gene,"human","mouse"),
                                         human_prop=gene_human_prop,
                                         mouse_prop=gene_mouse_prop)
    
    wrong_prop[1,dset] <- 1-sum(raw_mat[human_gene,human_spots])/
        sum(raw_mat[human_gene,])
    wrong_prop[2,dset] <- 1-sum(raw_mat[!human_gene,mouse_spots])/
        sum(raw_mat[!human_gene,])
    
    wrong_prop2[1,dset] <- 1-sum(raw_mat[human_gene,human_spots])/
        sum(raw_mat[human_gene,c(human_spots,mouse_spots)])
    wrong_prop2[2,dset] <- 1-sum(raw_mat[!human_gene,mouse_spots])/
        sum(raw_mat[!human_gene,c(human_spots,mouse_spots)])
    
    LPSS[dset] <- (sum(raw_mat[human_gene,slide_tmp$species%in%c("mouse","background")])+
        sum(raw_mat[!human_gene,slide_tmp$species%in%c("human","background")]))/sum(raw_mat)
    
    BG[dset] <- sum(raw_mat[,slide_tmp$tissue==0])/sum(raw_mat)
}

# LPSS
round(LPSS,3)

# Proportion of background UMIs
round(BG,3)
```


Species-specific proportions in spots in raw data:

```{r, echo=F, dpi=300, fig.width=14, fig.height=3}
gp_list <- lpss_list <- list()
for(dset in names(slide_obj_list)){
    
    prop_df <- spot_prop_list[[dset]] %>% filter(species%in%c("human","mouse","background")) %>%
        melt(id.vars="species", measure.vars=c("human_prop","mouse_prop"))
    # prop_df <- spot_prop_list[[dset]] %>% #filter(species%in%c("human","mouse")) %>%
    #     melt(id.vars="species", measure.vars=c("human_prop2","mouse_prop2"))
    prop_df$hm_in_region <- paste(prop_df$variable, prop_df$species,sep = "_")
    
    
    
    gp <- ggplot(prop_df, aes(y=value, color=hm_in_region))+
        geom_boxplot()+ylab("Proportion")+ 
        ylim(c(0,1))+
        # ylim(c(0,2e-3))+
        theme_bw()+
        ggtitle(NULL)+
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              panel.grid.major.x = element_blank(), 
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"),legend.position = "none")
    
    gp_list[[dset]] <- gp
    
    
    prop_df <- spot_prop_list[[dset]] %>% filter(species%in%c("human","mouse","background")) %>% 
        select(species,lpss)
    
    gp <- ggplot(prop_df, aes(y=lpss, color=species))+
        geom_boxplot()+labs(y="LPSS",color="spots")+theme_bw()+ylim(c(0,5e-4))+
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"))+ggtitle(dset)
    
    lpss_list[[dset]] <- gp

}
# tiff(file.path(tif_dir,"MS_down_hmprop_nol.tif"),
#      res = 300, units = "in", width = 10,height = 3)
plot_grid(plotlist = gp_list[c(2,1,3)], align = "vh", ncol=3)
# plot_grid(plotlist = lpss_list[c(2,1,3)], align = "vh", ncol=3)
# dev.off()
```

### Run decontamination

Due to low expression, set a smaller tolerance.

```{r, echo=F, eval=F}
decont_obj_list <- list()

for(dset in names(slide_obj_list)){
    decont_obj_list[[dset]] <- SpotClean(slide_obj_list[[dset]],
                                         gene_keep = rownames(slide_obj_list[[dset]]),tol = 0.1)
}

save(decont_obj_list, file="~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/mixed_species_decont_down.RData")

```

### Calculate summary statistics after decontamination

```{r, echo=F}
load("~/Google Drive/Hallu/codes/ckgroup/spatial_data/manuscript/codes/mixed_species_decont_down.RData")
tissue_spot_prop_list <- lapply(spot_prop_list, function(x) filter(x, species!="background"))

wrong_prop_decont <- wrong_prop

for(dset in names(decont_obj_list)){
    decont_mat <- assay(decont_obj_list[[dset]])
    slide_tmp <- metadata(slide_obj_list[[dset]])$slide %>% filter(tissue==1)
    
    human_gene <- grepl("GRCh38",rownames(decont_mat))
    
    spot_human_umi <- colSums(decont_mat[human_gene,])
    spot_mouse_umi <- colSums(decont_mat[!human_gene,])
    
    spot_human_prop <- spot_human_umi/(spot_human_umi+spot_mouse_umi)
    spot_human_prop2 <- spot_human_umi/sum(decont_mat)
    spot_mouse_prop2 <- spot_mouse_umi/sum(decont_mat)
    
    tissue_spot_prop_list[[dset]]$decont_human_counts <- spot_human_umi
    tissue_spot_prop_list[[dset]]$decont_mouse_counts <- spot_mouse_umi
    
    tissue_spot_prop_list[[dset]]$decont_human_prop <- spot_human_prop
    tissue_spot_prop_list[[dset]]$decont_mouse_prop <- 1-spot_human_prop
    
    tissue_spot_prop_list[[dset]]$decont_human_prop2 <- spot_human_prop2
    tissue_spot_prop_list[[dset]]$decont_mouse_prop2 <- spot_mouse_prop2
    
    gene_prop_list[[dset]]$decont_wrong_prop <- 
        gene_prop_list[[dset]]$wrong_prop <- NA
    
    h_spots <- tissue_spot_prop_list[[dset]]$species=="human"
    m_spots <- tissue_spot_prop_list[[dset]]$species=="mouse"
    
    # estimated contamination rates
    tissue_spot_prop_list[[dset]]$cont_rate <- metadata(decont_obj_list[[dset]])$contamination_rate
    
    # Per-spot LPSS after decontamination
    lpss <- numeric(nrow(slide_tmp))
    lpss[h_spots] <- colSums(decont_mat[!human_gene,h_spots])/sum(decont_mat)
    lpss[m_spots] <- colSums(decont_mat[human_gene,m_spots])/sum(decont_mat)

    tissue_spot_prop_list[[dset]]$decont_lpss <- lpss
    
    # proportion of UMIs in human spots for each gene
    gene_human_prop <- rowSums(decont_mat[,h_spots])/rowSums(decont_mat)
    
    # proportion of UMIs in mouse spots for each gene
    gene_mouse_prop <- rowSums(decont_mat[,m_spots])/rowSums(decont_mat)
    
    gene_prop_list[[dset]]$decont_human_prop <- gene_human_prop
    gene_prop_list[[dset]]$decont_mouse_prop <- gene_mouse_prop
    
    wrong_prop_decont[1,dset] <- sum(decont_mat[human_gene,m_spots])/
        sum(decont_mat[human_gene,])
    wrong_prop_decont[2,dset] <- sum(decont_mat[!human_gene,h_spots])/
        sum(decont_mat[!human_gene,])
    
    # wrong prop per gene
    gene_prop_list[[dset]]$decont_wrong_prop[ human_gene] <- 
        1- gene_prop_list[[dset]]$decont_human_prop[ human_gene]
    gene_prop_list[[dset]]$decont_wrong_prop[!human_gene] <- 
        1-gene_prop_list[[dset]]$decont_mouse_prop[! human_gene]
    
    gene_prop_list[[dset]]$wrong_prop[ human_gene] <- 
        1-gene_prop_list[[dset]]$human_prop[ human_gene]
    gene_prop_list[[dset]]$wrong_prop[! human_gene] <- 
        1-gene_prop_list[[dset]]$mouse_prop[! human_gene]
}

wrong_prop %>% round(3)
wrong_prop_decont %>% round(3)

total_df <- data.frame(species=rep(c("human","mouse"),2),method=rep(c("raw","SpotClean"),each=2),val=c(wrong_prop[,2],wrong_prop_decont[,2]))
# tiff(file.path(tif_dir,"MS_1_wrong_prop.tif"),units = "in", res = 300, width = 2,height = 3.3)
ggplot(total_df, aes(x=species,y=val,group=method, fill=method))+geom_bar(stat="identity",position = "dodge")+
    labs(fill=NULL, y="proportion of spot-swapped UMIs")+
    theme(axis.text.x=element_blank(),
          axis.title.x=element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid = element_blank(),legend.position = "top")
# dev.off()
```

### Species-specific proportions in spots in decont data

```{r, echo=F, dpi=300, fig.width=14, fig.height=3}
gp_list <- lpss_list <- list()
for(dset in names(slide_obj_list)){
    
    prop_df <- tissue_spot_prop_list[[dset]] %>% filter(species%in%c("human","mouse")) %>%
        melt(id.vars="species", measure.vars=c("decont_human_prop","decont_mouse_prop"))

    prop_df$hm_in_region <- paste(prop_df$variable, prop_df$species,sep = "_")
    
    gp <- ggplot(prop_df, aes(y=value, color=hm_in_region))+
        geom_boxplot()+ylab("Proportion")+ 
        ylim(c(0,2e-3))+
        theme_bw()+
        ggtitle(dset)+
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"))
    
    gp_list[[dset]] <- gp
    
    gp <- ggplot(tissue_spot_prop_list[[dset]], aes(y=lpss, color=species))+
        geom_boxplot()+labs(y="LPSS",color="spots")+theme_bw()+ylim(c(0,5e-4))+
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "black"))+ggtitle(dset)
    
    lpss_list[[dset]] <- gp
    
}

# tiff(file.path(tif_dir,"MS_raw_hmprop_l.tif"),
#      res = 300, units = "in", width = 10,height = 3)
plot_grid(plotlist = gp_list[c(2,1,3)], align = "vh", ncol=3)
# dev.off()
```

### Plot spot and gene proportions (species purity) in top human and mouse spots and genes

```{r, echo=F, dpi=300, fig.width=14, fig.height=5}
s_top_n <- 100
g_top_n <- 100

for(dset in names(slide_obj_list)){
    
    ###################
    # genes
    ##################
    
    human_top <- gene_prop_list[[dset]] %>% filter(species=="human") %>% 
        arrange(-ave_exp) %>% head(g_top_n)

    mouse_top <- gene_prop_list[[dset]] %>% filter(species=="mouse") %>% 
        arrange(-ave_exp) %>% head(g_top_n)
    
    h_range <- range(1-human_top$human_prop,1-human_top$decont_human_prop)
    m_range <- range(1-mouse_top$mouse_prop,1-mouse_top$decont_mouse_prop)
    
    g1 <- ggplot(human_top %>% filter(species=="human"), 
                 aes(x=1-human_prop, y=1-decont_human_prop))+
        geom_abline(slope = 1, color="pink",size=1)+
        geom_point()+lims(x=h_range,
                          y=h_range)+theme_bw(base_rect_size = 2)+
        labs(#title=paste0(dset,": top 100 human genes"),
             x="raw\n",
             y="\nSpotClean")+
        theme(axis.title = element_text(size = 20),
              axis.text = element_text(size=15),
              panel.grid.minor = element_blank(),
              plot.margin = margin(t=0.1,r = 0.2, unit = "in"))
    
    g2 <- ggplot(mouse_top %>% filter(species=="mouse"), 
                 aes(x=1-mouse_prop, y=1-decont_mouse_prop))+
        geom_abline(slope = 1, color="pink",size=1)+
        geom_point()+lims(x=m_range,
                          y=m_range)+theme_bw(base_rect_size = 2)+
        labs(#title=paste0(dset,": top 100 mouse genes"),
             x="raw\n",
             y="\nSpotClean")+
        theme(axis.title = element_text(size = 20),
              axis.text = element_text(size=15),
              panel.grid.minor = element_blank(),
              plot.margin = margin(t=0.1,r = 0.2, unit = "in"))
    
    
    ###################
    # spots
    ###################
    
    human_top <- tissue_spot_prop_list[[dset]] %>% filter(species=="human") %>% 
        arrange(-decont_human_counts) %>% head(s_top_n)
    
    mouse_top <- tissue_spot_prop_list[[dset]] %>% filter(species=="mouse") %>% 
        arrange(-decont_human_counts) %>% head(s_top_n)
    
    h_range <- range(1-human_top$human_prop,1-human_top$decont_human_prop)
    m_range <- range(1-mouse_top$mouse_prop,1-mouse_top$decont_mouse_prop)
    
    s1 <- ggplot(human_top, 
                 aes(x=1-human_prop, y=1-decont_human_prop))+
        geom_abline(slope = 1, color="pink",size=1)+
        geom_point()+lims(x=h_range,
                          y=h_range)+theme_bw(base_rect_size = 2)+
        labs(#title=paste0(dset,": top 100 human spots"),
             x="raw\n",
             y="\nSpotClean")+
        theme(axis.title = element_text(size = 20),
              axis.text = element_text(size=15),
              panel.grid.minor = element_blank(),
              plot.margin = margin(t=0.1,r = 0.2, unit = "in"))
    
    s2 <- ggplot(mouse_top, 
                 aes(x=1-mouse_prop, y=1-decont_mouse_prop))+
        geom_abline(slope = 1, color="pink",size=1)+
        geom_point()+lims(x=m_range,
                          y=m_range)+theme_bw(base_rect_size = 2)+
        labs(#title=paste0(dset,": top 100 mouse spots"),
             x="raw\n",
             y="\nSpotClean")+
        theme(axis.title = element_text(size = 20),
              axis.text = element_text(size=15),
              panel.grid.minor = element_blank(),
              plot.margin = margin(t=0.1,r = 0.2, unit = "in"))
    
    tiff(file.path(tif_dir,paste0(dset,"_",g_top_n,"g",s_top_n,"s_wrongprop.tif")),res = 300,
         width = 8, height = 7, units = "in")
    plot(plot_grid(g1, g2, s1, s2, align = "vh", ncol = 2))
    dev.off()

}


```

Why small spots look not good: for small spots, the proportion is not necessarily improved even correctly decontaminated.

Example: an empty human spot received the same amount of human and mouse contamination. After decontamination, more human contamination was removed than mouse contamination. As a result, both human and mouse contamination were reduced, but human proportion (species purity) decreased.

Why large spots are better: large spots are unlikely to be empty spots, and its majority expression comes from the correct species. After decontamination, it will receive expression of the correct species and lose expression of the wrong species, thus improves species purity.

Why large genes are preferred: lowly expressed genes are sparse. SpotClean does not change too much of sparse genes.

### Plot wrong prop in human and mouse genes before and after decontamination

```{r, echo=F}
g_top_n <- 5000

for(dset in names(slide_obj_list)){
    
    human_top <- gene_prop_list[[dset]] %>% filter(species=="human") %>% 
        arrange(-ave_exp) %>% head(g_top_n)
    
    mouse_top <- gene_prop_list[[dset]] %>% filter(species=="mouse") %>% 
        arrange(-ave_exp) %>% head(g_top_n)
    
    plot(ggplot(rbind(human_top, mouse_top),
           aes(x=wrong_prop, y=decont_wrong_prop, color=species))+
        geom_point()+
        geom_abline(slope = 1, color="pink"))
    
}

```

### For top 100 human (mouse) spots, check gene expression distributions

Check top 20 human and top 20 mouse gene expression before and after decontamination. These selected genes are top among the whole data, not just in these spots.

For better visualization, the largest proportions in the heatmaps were trimmed to 80% quantile.

```{r, echo=F}
g_top_n <- 100
s_top_n <- 100

spot_exp_list <- list()

for(dset in names(slide_obj_list)){
    # raw and decont matrix
    mat_list <- list(raw=assay(slide_obj_list[[dset]])[
        ,metadata(slide_obj_list[[dset]])$slide$tissue==1
    ],
    decont=assay(decont_obj_list[[dset]])
    )
    
    slide_tmp <- decont_obj_list[[dset]]@metadata$slide
    
    human_s <- colSums(
        mat_list$decont[,filter(slide_tmp,species=="human")$barcode]
    ) %>% sort(decreasing = T) %>% head(s_top_n) %>% names
    mouse_s <- colSums(
        mat_list$decont[,filter(slide_tmp,species=="mouse")$barcode]
    ) %>% sort(decreasing = T) %>% head(s_top_n) %>% names
    
    human_g <- rowSums(
        mat_list$decont[grepl("GRCh38",rownames(mat_list$decont)),]
    ) %>% sort(decreasing = T) %>% head(g_top_n) %>% names
    mouse_g <- rowSums(
        mat_list$decont[!grepl("GRCh38",rownames(mat_list$decont)),]
    ) %>% sort(decreasing = T) %>% head(g_top_n) %>% names
    
    # reorder spots by hierarchical clustering
    hc <- hclust(as.dist(1-cor(as.matrix(mat_list$decont[,human_s]))))
    human_s <- hc$labels[hc$order]
    
    hc <- hclust(as.dist(1-cor(as.matrix(mat_list$decont[,mouse_s]))))
    mouse_s <- hc$labels[hc$order]
    
    # calculate proportions
    spot_exp_list[[dset]] <- list()
    for(meth in names(mat_list)){
        
        spot_exp_list[[dset]][[meth]] <- (
            (mat_list[[meth]][,c(human_s,mouse_s)])/
                rowSums(mat_list[[meth]][,c(human_s,mouse_s)])
        )[c(human_g,mouse_g),]*1e4
    }
}

```

```{r, echo=F, dpi=300, fig.width=8, fig.height=3}
myPalette <- grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(11, "Spectral")))
log_idx <- log(1:100)/sum(log(1:100))
log_idx <- round(log_idx/max(log_idx)*1000)+1
cols <- myPalette(100)

# truncate top 80% for better visualization
q_trim <- 0.8
dp_list <- list()
for(dset in names(slide_obj_list)){
    
    # color scale range
    max_prop <- lapply(spot_exp_list[[dset]], function(x) quantile(as.matrix(x), q_trim)) %>% unlist %>% max
    min_prop <- min(spot_exp_list[[dset]]$raw@x,spot_exp_list[[dset]]$decont@x)
    
    breaks <- expm1(log1p(min_prop)+ (log1p(max_prop)-log1p(min_prop))/4*0:4 )    
    
    dp_tmp <- list()
    for(meth in c("raw","decont")){
        
        exp_df <- melt(spot_exp_list[[dset]][[meth]] %>% as.matrix,varnames=c("gene","barcode"))
        exp_df$value <- pmin(exp_df$value,quantile(exp_df$value,q_trim))
        
        gp <- ggplot(exp_df,aes(x=barcode, y=gene, fill=value))+geom_tile()+
            scale_fill_gradientn(colours = cols,limits=c(0,max_prop), trans="log1p", 
                                 breaks=floor(breaks),labels=floor(breaks)/1e4)+
            theme(axis.text = element_blank(),
                  axis.ticks = element_blank(),
                  axis.title = element_blank())+
            labs(fill="scaled UMIs")+
            geom_vline(xintercept = s_top_n+0.5)+
            geom_hline(yintercept = g_top_n+0.5)
        
        dp_list[[paste(meth,dset)]] <- gp
        dp_tmp[[meth]] <- gp
    }

    # tiff(file.path(tif_dir,paste0(dset,"_",g_top_n,"g",s_top_n,"s","_heatmap.tif")),
    #      res = 300, units = "in",width = 6, height = 2.8)
    plot(plot_grid(plotlist = dp_tmp, ncol = 2, align = "vh"))
    # dev.off()
}

```

### Visualize top 5 expressed genes

For better visualization, expressions less than 1 were trimmed out.

Mouse genes in all spots: 

```{r, echo=F, dpi=300, fig.width=11, fig.height=4, warning=F}
top_n <- 5
for(dset in names(slide_obj_list)){

    slide_tmp <- metadata(decont_obj_list[[dset]])$slide
    hm_bcs <- filter(slide_tmp, species%in%c("human","mouse"))$barcode
    genes <- top1000_list[[dset]]$mouse[seq_len(top_n)]
    
    dp_list <- list()
    for(gn in genes){
        
        target_range <- c(1,max(assay(decont_obj_list[[dset]])[gn,hm_bcs]))
        
        dp1 <- VisualizeHeatmap(slide_obj_list[[dset]],gn, title = NULL, 
                                subset_barcodes = hm_bcs,
                                legend_title = gsub(".*_","",gn),
                                legend_range = target_range)+
            labs(subtitle=m_gene_prop[gn,"mouse_prop"])#+.theme_void()
        
        dp2 <- VisualizeHeatmap(decont_obj_list[[dset]],gn, title = NULL, 
                                subset_barcodes = hm_bcs,
                                legend_title = gsub(".*_","",gn),
                                legend_range = target_range)+
            labs(subtitle=m_gene_prop[gn,"decont_mouse_prop"])#+.theme_void()
        
        plot(plot_grid(dp1, dp2, ncol = 2, align = "vh"))
        
        dp_list <- c(dp_list,list(dp1, dp2))
    }
    
    #tiff(file.path(tif_dir,paste0(dset,"_mouse_decont_ l.tif")),res = 300, units = "in", width = 11, height = 22)
    #plot(plot_grid(plotlist = dp_list,  ncol = 2, align = "vh"))
    #dev.off()
}

```

Human genes in all spots:

```{r, echo=F, dpi=300, fig.width=11, fig.height=4, warning=F}
for(dset in names(slide_obj_list)){

    slide_tmp <- metadata(decont_obj_list[[dset]])$slide
    hm_bcs <- filter(slide_tmp, species%in%c("human","mouse"))$barcode
    genes <- top1000_list[[dset]]$human[seq_len(top_n)]
    
    dp_list <- list()
    for(gn in genes){
        

        target_range <- c(1,max(assay(decont_obj_list[[dset]])[gn,hm_bcs]))
        
        
        dp1 <- VisualizeHeatmap(slide_obj_list[[dset]],gn, title = NULL, 
                                subset_barcodes = hm_bcs,
                                legend_title = gsub(".*_","",gn),
                                legend_range = target_range)+.theme_void()
        
        dp2 <- VisualizeHeatmap(decont_obj_list[[dset]],gn, title = NULL, 
                                subset_barcodes = hm_bcs,
                                legend_title = gsub(".*_","",gn),
                                legend_range = target_range)+.theme_void()
        
        plot(plot_grid(dp1, dp2, ncol = 2, align = "vh"))
        
        dp_list <- c(dp_list,list(dp1, dp2))
    }

    #tiff(file.path(tif_dir,paste0(dset,"_human_decont_nol.tif")),res = 300, units = "in", width = 11, height = 22)
    # plot(plot_grid(plotlist = dp_list,  ncol = 2, align = "vh"))
    #dev.off()
}

```

